---
title: "Data Formatting and Manipulations"
output: html_notebook
---

***

## Overview
Reading in the data as batches in [Project and Data Setup](http://htmlpreview.github.io/?https://github.com/chris-w-r/lichess-database-exploration/blob/master/Project%20and%20Data%20Setup.nb.html) saves all of the values in the columns as `strings`.  While it may be helpful for the future to retain copies of the data in their raw formats as received from [Lichess](https://lichess.org/), we want to format and manipulate these values into variables types which R prefers -- `numerics` and `factors`.

We will format and manipulate each column (or group of similar columns such as titles) leveraging [Wikipedia's PGN article](https://en.wikipedia.org/wiki/Portable_Game_Notation) as a general guide to expected values.  If a column requires a specific reference guide, it is included in its section.

***

## Code

* [Setup and Load Data](#setup)
* [Event](#event)
* [Result](#result)
* [Elo](#elo)
* [RatingDiff](#ratingdiff)
* [Title](#title)
* [ECO](#ECO)
* [Time Control](#timecontrol)
* [Termination](#termination)
* [Movetext](#movetext)

### Setup and Load Data {#setup}
```{r setup}
knitr::opts_chunk$set(eval = FALSE)
```


```{r load data}
library(data.table)
library(tidyverse)

drop_columns_after_manipulation <- FALSE
check_factors_lost_data <- TRUE


load(file = paste0(lichess_db_standard_rated_filepath,
                   "april_eval_base.rda"))
april_eval_base <- april_eval_base[!is.na(Event)]


# load(file = paste0(lichess_db_standard_rated_eval_stage_filepath,
#                    "april_1.rda"))
# april_eval_base <- april_1
```

### Event {#event}
```{r event}
april_eval_base[, TimeControlVariation := factor(
  sub("^\\w+ (\\w+).+", "\\1", Event),
  levels = c(
  "UltraBullet",
  "Bullet",
  "Blitz",
  "Classical",
  "Correspondence"
  )
  )][, Tournament := factor(
  ifelse(
  sub("^\\w+ \\w+ (\\w+).*", "\\1", Event) == "game",
  "Non-Tournament",
  "Tournament"
  ),
  levels = c("Non-Tournament", "Tournament")
  )][, TournamentSite := ifelse(grepl("game", Event),
  NA,
  sub("^\\w+ \\w+ \\w+ (.*)", "\\1", Event))]
  
if (check_factors_lost_data)
{
  event_original <- as.data.table(april_eval_base$Event)
  
  if (nrow(event_original[!grepl("^Rated", V1)]) > 0)
  {
    stop("First word other than 'Rated' found!")
  }
  
  event_original[, TimeControlVariation := sub("^\\w+ (\\w+).+", "\\1", V1)]
  if (nrow(count(april_eval_base[!is.na(TimeControlVariation)], TimeControlVariation)) != nrow(count(event_original, TimeControlVariation)))
  {
    stop("Additional TimeControlVariation found!")
  }
  
  event_original[, Tournament := sub("^\\w+ \\w+ (\\w+).*", "\\1", V1)]
  if (nrow(count(april_eval_base[!is.na(Tournament)], Tournament)) != nrow(count(event_original, Tournament)))
  {
    stop("Third word other than 'game' or 'tournament' found!")
  }
  
  rm(event_original)
}
  
if (drop_columns_after_manipulation)
{
  april_eval_base[, Event := NULL]
}
```

### Result {#result}
```{r result}
april_eval_base[, Winner := factor(ifelse(grepl("[*]", Result), "Abandoned", ifelse(grepl("2", Result), "Draw", ifelse(grepl("^1", Result), "White", "Black"))), levels = c("White", "Black", "Draw", "Abandoned"))]

if(check_factors_lost_data)
{
  if(nrow(count(april_eval_base[!is.na(Winner)], Winner)) != nrow(count(april_eval_base, Result)))
  {
    stop("Additional Result found!")
  }
}

if(drop_columns_after_manipulation)
{
  april_eval_base[, Result := NULL]
}
```

### Elo {#elo}
<https://en.wikipedia.org/wiki/Chess_rating_system#Elo_rating_system>
```{r elo}
april_eval_base$WhiteElo <- as.numeric(april_eval_base$WhiteElo)
april_eval_base$BlackElo <- as.numeric(april_eval_base$BlackElo)

elo_breaks <- c(
  -Inf,
  1200,
  1400,
  1600,
  1800,
  2000,
  2200,
  2300,
  2400,
  2500,
  2700,
  Inf
)
elo_labels <- c(
  "Below 1200",
  "1200 - 1399",
  "1400 - 1599",
  "1600 - 1799",
  "1800 - 1999",
  "2000 - 2199",
  "2200 - 2299",
  "2300 - 2399",
  "2400 - 2499",
  "2500 - 2699",
  "2700+"
  )
april_eval_base$WhiteEloRange <- cut(april_eval_base$WhiteElo,
                                     breaks = elo_breaks,
                                     labels = elo_labels, 
                                     right = FALSE,
                                     ordered_result = TRUE)
april_eval_base$BlackEloRange <- cut(april_eval_base$BlackElo,
                                     breaks = elo_breaks,
                                     labels = elo_labels, 
                                     right = FALSE,
                                     ordered_result = TRUE)
rm(elo_breaks)
rm(elo_labels)
```

### RatingDiff {#ratingdiff}
```{r ratingdiff}
april_eval_base$WhiteRatingDiff <- as.numeric(april_eval_base$WhiteRatingDiff)
april_eval_base$BlackRatingDiff <- as.numeric(april_eval_base$BlackRatingDiff)
```

### Title {#title}
<https://lichess.org/help/master>
```{r title}
if(check_factors_lost_data)
{
  whitetitle_original <- as.data.table(april_eval_base$WhiteTitle)
  blacktitle_original <- as.data.table(april_eval_base$BlackTitle)
}

title_levels <-
  c(NA,
    "LM",
    "WNM",
    "NM",
    "WCM",
    "CM",
    "WFM",
    "FM",
    "WIM",
    "IM",
    "WGM",
    "GM")
april_eval_base$WhiteTitle <- factor(april_eval_base$WhiteTitle,
                                     levels = title_levels)
april_eval_base$BlackTitle <- factor(april_eval_base$BlackTitle,
                                     levels = title_levels)
rm(title_levels)

if(check_factors_lost_data)
{
  if(nrow(count(april_eval_base[!is.na(WhiteTitle)], WhiteTitle)) != nrow(count(whitetitle_original[!is.na(V1)], V1)))
  {
    stop("Additional WhiteTitle found!")
  }
  if(nrow(count(april_eval_base[!is.na(BlackTitle)], BlackTitle)) != nrow(count(blacktitle_original[!is.na(V1)], V1)))
  {
    stop("Additional BlackTitle found!")
  }
  
  rm(whitetitle_original)
  rm(blacktitle_original)
}
```

### ECO {#ECO}
```{r ECO}
#TBD
```

### Time Control {#timecontrol}
```{r time control}
april_eval_base[, TimeControlMinutes := ifelse(grepl("[-]", TimeControl), NA, as.numeric(sub("(\\d+).+", "\\1", TimeControl)) / 60)][, TimeControlIncrement := ifelse(grepl("[-]", TimeControl), NA, as.numeric(sub("\\d+[+](\\d+)", "\\1", TimeControl)))]

if(drop_columns_after_manipulation)
{
  april_eval_base[, TimeControl := NULL]
}
```

### Termination {#termination}
```{r termination}
if(check_factors_lost_data)
{
  termination_original <- as.data.table(april_eval_base$Termination)
}

april_eval_base$Termination <- factor(april_eval_base$Termination, levels = c("Normal", "Time forfeit", "Abandoned", "Rules infraction"))

if(check_factors_lost_data)
{
  if(nrow(count(april_eval_base[!is.na(Termination)], Termination)) != nrow(count(termination_original, V1)))
  {
    stop("Additional Termination found!")
  }
  
  rm(termination_original)
}
```

### Movetext {#movetext}
```{r checkmate flag}
april_eval_base[, Checkmate := grepl("# ", Movetext)]
```


```{r movetext}

library(dplyr)
library(purrr)
library(stringr)

format_moves <- function(movetext)
{
  movetext_df <-
    str_match_all(
      movetext,
      "(\\d+)([.]+) ([a-h1-8KQRBNxO=#+-]+)([?!]*) \\{( \\[%eval ([0-9.#-]+)\\])? \\[%clk ([0-9:]+)\\] \\}"
    )
  
  movetext_df <- as.data.frame(movetext_df[[1]][, c(-1, -6)])
  
  colnames(movetext_df) <-
    c(
      "MoveNumber",
      "Indicator",
      "AlgebraicNotation",
      "Annotation",
      "Evaluation",
      "Clock"
    )
  
  movetext_df
}

batch_read <- batch_read %>%
    mutate(Moves = map(Moves, format_moves))

```

